Men and past smokers from ontario are potentially more physically active than women and non-smokers

### 2024-12-02 paper/paper.qmd
Major update for paper
1. cleaned up paper.qmd file:
 - added diagnostic check in appendix
 - added one more discussion point

2. removed unnecessary files

3. added to model justification regarding BIC values

4. added parquet file for analysis data

5. cleaned references

---
title: "Men and past smokers from ontario are potentially more physically active than women and non-smokers"
subtitle: "Women are engaging in lesser physical activities than men. Furthermore, individuals who smoked before are more likely to engage in exercise. Consumption of stimulant-related drugs could have also increased as well, resulting in increased overall physical activities in Ontario."
author: 
  - Chris Yong Hong Sen
thanks: "Code and data are available at: [https://github.com/Monoji77/Alcohol_Use_Ontario](https://github.com/Monoji77/Alcohol_Use_Ontario)."
date: today
date-format: long
abstract: "This study explores factors influencing physical activity levels in Ontario, focusing on demographic and substance abuse. Our findings reveal that women engage in less physical activity than men, while individuals who have smoked in the past and those consuming stimulant-related drugs are more likely to exercise. Additionally, age, highest educational attainment, and health region also play significant roles in determining physical activity levels. These insights highlight the complex interplay of gender, lifestyle behaviors, and sociodemographic factors, providing valuable information for designing targeted public health interventions to promote physical activity."
format: 
  pdf:
    fig-pos: H
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(palmerpenguins)
library(knitr)
```

@@

Overview paragraph

Estimand paragraph
The estimand would be the actual effect of socio-demographic factors, substance abuse, and health factors on how much time is spent on strenuous activities among individuals in Ontario who engage in regular exercise.

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

Other datasets explored include sources from World Health Organisation and ParticipACTION. World Health Organisation provides aggregated data informing certain indicators of interest such as obesity levels and alcohol consumption. This can be found in the data section for Canada[@OtherSource1]. Also, ParticipACTION provides key statistics regarding proportion of adult population in canada meeting the national guidelines of 150 minutes of moderate to vigorous exercise, as well as how sedentary canadians are [@OtherSource2]. However, the availability of raw individual level data is absent from these sources. Although the aggregated data gives us a sensing about general Canadian health, we are not able to focus our attention to the Ontario population. The dataset chosen from Canadian Community Health Survey (CCHS) annual component is ideal as it contains individual level data, where we are able to obtain the province the respondents are from.

@@

Our goal is to explore the actual effect of socio-demographic factors, substance abuse, and health factors on how much time is spent on exercise among individuals in Ontario who engage in regular exercise. The instrument of measurement would be surveys. Through the cross-sectional survey conducted by CCHS, we are able to get the number of hours spent on vigorous exercise per individual, our response variable.

According to the questionnaire, respondents were asked in a phone survey to indicate whether they had engaged in any strenuous activity that lasted at least 10 minutes and made them sweat and breathe a little harder in the past 7 days. Afterwards, they were asked to indicate the total time spent on those activities. To address the concern of recall bias where an individual may not recall the specific timing in engaging with exercise, surveyors sectioned vigorous exercise into commuting, sports, work and volunteering. Then within each section, surveyors asked which day (Monday to Sunday) in the past week had they engaged in such exercise that made them sweat and breathe a little harder. By doing so, the activities individuals did during the past week would be anchored on the days of the week. Following this would be for the respondents to indicate the total number of hours they engaged in each
section of vigorous exercise for the past 7 days.

@@

For socio-demographic factors possibly affecting time spent on vigorous exercise, the surveyors asked the respondent for their age (DHH_AGE) from 0 years old to 121 years old), sex (DHH_SEX) being either male or female, personal income obtained from tax forms submitted to the Canada Revenue Agency upon consent from respondent, geographical health region (ADM_Q037) that was provided by respondents when they agreed to the phone survey, and highest educational attainment (EHG2_04) ranging from less than high school diploma to master's degree and above.

@@

Overall, each entry obtained by respondents assigns quantitative values to each factor and the the response for which we can explore subsequently using linear regression to determine which factors have greater influence in the time spent exercising amongst individuals that are engaged in exercising.

## Response variable: Time spent on vigorous activity in the past 7 days

From the CCHS dataset, we look at paadvmva variable. This variable is continuous with values ranging from 0 to 9902 representing the number of minutes spent on vigorous activity in the past 7 days. For easier readability, the variable will be renamed as time_spent_vigorous_exercise_7d.  As explained in section @sec-measurement, this variable measures the time spent on any activities lasting more than 10 minutes that makes an individual sweat more than they normally would. This would include individuals undertaking laborious employment or volunteering efforts, or it could simply include individuals who do cardio and/or strength training.

By using this variable, we are assuming that the routine of an individual from the past 7 days is indicative of their general routine beyond the past 7 days. This variable may not account for individuals who only began vigorous sweating routine within the past 7 days and were sedentary before, which may not be helpful if included in subsequent model fitting.

Since we are interested in individuals that are already engaging in strenuous activities, we can filter time_spent_vigorous_exercise_7d to exclude 0. After removing such entries, we obtain the following summary statistics as shown in @tbl-summarised_statistics_time_spent_vigorous_exercise_7d. The median time spent on vigorous activities the past 7 days would be 300 minutes, which averages out to 43 minutes per day.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summarised_statistics_time_spent_vigorous_exercise_7d
#| tbl-cap: Summarised statistics for time spent on vigorous exercise in the past 7 days

original_data <- read_csv("../data/02-analysis_data/analysis_data.csv")

@@

kable(data_cleaned_1 |>
  summarise(`Min.` = min(time_spent_vigorous_exercise_7d),
            `1st Qu.` = quantile(time_spent_vigorous_exercise_7d, 0.25),
            Median = median(time_spent_vigorous_exercise_7d),
            `3rd Qu.` = quantile(time_spent_vigorous_exercise_7d, 0.75),
            `Max.` = max(time_spent_vigorous_exercise_7d)))

@@

Interestingly, the maximum time spent on vigorous activities the past 7 days was 9902 minutes, and this averages to 1415 minutes per day, which is equivalent to 23.5 hours in a day. This is very unlikely as this suggests that an individual is constantly engaging in physical strenuous activity without sleeping. Logically, this suggests the possibility of outlier values. By observing @fig-boxplot_time_spent_vigorous_exercise_7d, there are many outlier observations for this variable. Roughly 10% of the observed data are considered outliers as they fall outside 1.5 times the interquartile range from the 25% and 75% quartile values. We would be performing log transformation which would be discussed in section (INSERT SECTION HERE PLEASE DONT FORGET) which would address these outlier values.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-boxplot_time_spent_vigorous_exercise_7d
#| fig-cap: Boxplot of time spent on vigorous exercise in the past 7 days
#boxplot(data_cleaned_1$time_spent_vigorous_exercise_7d, outcol='indianred4')
ggplot(data_cleaned_1) +
  geom_boxplot(aes(y=time_spent_vigorous_exercise_7d), 
               outlier.color = 'indianred4', 
               outlier.alpha=0.2) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  ylab('') +
  xlab('Boxplot of time spent on vigorous exercise in the past 7 days')
```

Subsequently, we would have to remove observations that are very unlikely as they do not represent the general population and could skew the distribution of time spent on vigorous exercise.

@@

```{r response variable exploration}
#| echo: false
#| warning: false
#| message: false
#| label: fig-hist-time_spent_vigorous_exercise_7d
#| fig-cap: Histogram of time spent on vigorous exercise the past 7 days

hist(data_cleaned_1$time_spent_vigorous_exercise_7d,
     main='',
     xlab = 'time spent on vigorous exercise the past 7 days / min',
     ylab = 'no. of individuals',
     freq=TRUE, cex.main=1, las = 1
    )
```
## Interested explanatory variables

There are a total of 13 predictors of interest that is seemingly related to the time spent in vigorous exercise an exercising individual gets.

@@

From @fig-barplot-age, we can immediately see that most observations fall within age groups 30 to 34 years old (bin 6), and lesser observations fall within age groups 70-74 years old (bin 14).
```{r age predictor barplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot-age
#| fig-cap: Histogram of time spent on vigorous exercise the past 7 days

@@

From @tbl-summarised_statistics_age and @fig-boxplot-age, we are able to see that the the median age bin belongs to age group 40-44 years old. We are able to see that there are no outlier values from @fig-boxplot-age, which means that the survey has obtained an acceptable number of respondents across all age bins. 
```{r age predictor summarised statistics}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summarised_statistics_age
#| tbl-cap: Summarised statistics of respondent ages
kable(data_cleaned_1 |>
  summarise(`Min.` = min(age),
            `1st Qu.` = quantile(age, 0.25),
            Median = median(age),
            `3rd Qu.` = quantile(age, 0.75),
            `Max.` = max(age)))
```
```{r age predictor boxplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-boxplot-age
#| fig-cap: Boxplot of respondent ages

ggplot(data_cleaned_1) +
  geom_boxplot(aes(y=age), 
               outlier.color = 'indianred4', 
               outlier.alpha=0.2) +
  theme_minimal() +
  scale_y_continuous(breaks = c(0:15)) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  ylab('respondent ages') +
  xlab('')
```

@@

Sex is a possible influential predictor for total time spent on vigorous activity. Historically, men are associated with exercise and involved in more labour intensive roles such as construction which involves heavy lifting. According to @womenExerciseWHO, women are less active than men by 5% since 2000. However, with the advent of female only establishments such as female gyms, and 32% increase in memberships at fitness and health clubs from 2010 to 2019, it would be interesting to see if the sex of an individual still affects the time spent on vigorous activites among physically active individuals.

@@

In 2023, @EducationAndActiveLifestyle has done a study to show that highest educational attainment may be a leading factor for physical activity. Therefore, it would be worth exploring the effects of highest levels of education with the time spent on vigorous activities among the Ontario residents.

@@

```{r proportion of respondents, group(highest_educational_attainment)}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-proportion_highest_educational_attainment
#| tbl-cap: Proportion of respondents grouped by highest educational level
kable(data_cleaned_1 |>
  count(highest_educational_attainment, name = 'proportion_of_respondents') |>
  mutate(proportion_of_respondents = paste0(round(proportion_of_respondents/nrow(data_cleaned_1)*100,0),'%'),
         highest_educational_attainment = case_when(
           highest_educational_attainment == 1 ~ 'less than sec sch grad',
           highest_educational_attainment == 2 ~ 'sec sch grad',
           highest_educational_attainment == 3 ~ 'post sec certification and above',
           .default = '')))

@@

Health region is a categorical variable which tells us the geographical location an individual belongs to defined by Provincial Ministries of Health. It could be possible that certain health regions have more funding that are projected towards increasing awareness towards having an active lifestyle. Therefore, it is worth exploring the relationship between health region and time spent on vigorous activities.

There are a total of 34 different health regions in the Ontario province. They are encoded using 5 digits. Kindly refer to the appendix (INSERT APPENDIX HERE PLEASE DONT FORGET) to refer to all the various health regions. @fig-map_health_region shows us the various health regions in Ontario. The blue regions shows us health regions for which lesser respondents belong to while red regions show us health regions for which more respondents belong to.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-map_health_region
#| fig-cap: Map of top five and bottom five occuring Ontario health 
library(tmap)

@@

- 1: No income or income loss
- 2: Less than $20,000 
- 3: \$20,000 to \$39,999
- 4: \$40,000 to \$59,999
- 5: \$60,000 to \$79,999
- 6: \$80,000 or more

According to @fig-barplot_personal_income, most of our respondents (25% of the sample) in Ontario are earning between \$20,000 to \$39,000 as part of their annual personal income. It is also worth mentioning that there are very few respondents who are not earning any income in that year. They account for less than 0.5% of the sample. In the 2018 annual wages report provided by @PersonalIncomeTaxFiling2018, the annual personal income reported from tax filing in Ontario that year was $39,510. Therefore, the sampling population's personal income data is representative of Ontario individuals.

@@

### Variable 7: Illicit drug use the past 12 months

This is a binary variable which indicates whether an individual has used drugs in the past 12 months prior to the interview. Drugs include marijuana, hashish, cocaine, amphetamines, speed, methamphetamines, crystal meth, ecstasy, hallucinogens, and sniffing glue, gasoline or other solvents. If they have attempted any of these drugs, this variable would be flagged as '1' and '0' otherwise.

It would be interesting to see if drug consumption has an effect on exercising. @PhysicalActivityAndSubstanceAbuse exerts that individuals who are more active tend to have lower substance abuse, suggesting a correlation between these two. Having this variable would be interesting to explore the relationship of an individuals drug usage and their time spent on exercise.

@@

kable(data_cleaned_1 |>
  count(illicit_drug_use, name = 'num_of_respondents') |>
  mutate(prop_of_population = paste0(round(num_of_respondents/nrow(data_cleaned_1)*100,0),'%'),
         illicit_drug_use = ifelse(illicit_drug_use == 0, 
                                   'never consumed drugs',
                                   'consumed at least once')))
```

@@

It is well known that excessive smoking has negative health impacts on individuals. A cross sectional study conducted by @SmokingAndPhysicalActiveness found that smokers are less likely to be physically active compared to non-smokers. Hence, it would be interesting to discover if there is an effect of smoking more than 100 cigarettes on the time spent on exercise.

@fig-barplot-cigarettes shows us that within the sample, the visual difference between the number of respondents who smoked more than 100 cigarettes (blue column) and respondents who smoked lesser than 100 cigarettes are not that big. From @tbl-barplot-cigarettes, 57% of respondents have smoked less than 100 cigarettes while the remaining 43% of respondents smoked more than 100 cigarettes. As the classes in this variable is almost split half,

@@

@fig-pyrmaid-cigarettes shows us cigarette consumption by age bins among out sample population. As respondent age increases from the first bin (18 to 19 year old) to the fourth bin (30 to 34 year old), we can see the number of respondents who have smoked more than 100 cigarettes in their lifetime increased from less than 125 respondents to more than 500 respondents (about 3 times more respondents between the two age groups).
```{r cigarette consumption pyramid figure}
#| echo: false
#| warning: false
#| message: false
#| label: fig-pyrmaid-cigarettes
#| tbl-cap: Pyramid of cigarettes consumption in respondent's lifetime by age groups

@@

# Model

The goal of our model is to explore which factors affect time spent on physical activities. Therefore a simple exploratory linear regression would be the chosen 
modelling algorithm to determine it. A full model where our response (time_spent_on_physical_activities) would be fitted on all our interested variables outlined in @sec-data-interestedvariables (INSERT THE SECTION  HERE LATER PLEASE DONT FORGET). Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

\begin{align*} 
y_k  &= \beta_0 + \beta_1Age_k + \beta_2Sex_k + \beta_3Educ_k + \beta_4Alc_k +\beta_5I*(Drug_k) + \beta_6I*(Cigar_k) + \\ 
     & \quad\: \beta_7I*(HealthRegion_k=35953) + \beta_8*I(HealthRegion_k=35970) + \epsilon_k
\end{align*}

\begin{align} 
y_k|\mu_k, \sigma^2 &\sim \mbox{Normal}(\mu_k, \sigma^2) \tag{1} \\
\mu_k &= \beta_0 + \beta_1Age_k + \beta_2Sex_k + \beta_3Educ_k + \beta_4Alc_k + \tag{2} \\
      & \quad\: \beta_5I*(Drug_k) + \beta_6I*(Cigar_k) + \notag \\ 
      & \quad\: \beta_7I*(HealthRegion_k=35953) + \notag \\
      & \quad\: \beta_8*I(HealthRegion_k=35970) \notag \\
\epsilon_k &\sim \mbox{Normal}(0, \sigma^2) \tag{3}  
\end{align}

@@

### Model justification

@@

\begin{align*} 
y_k  &= \beta_0 + \beta_1Age_k + \beta_2Sex_k + \beta_3Educ_k + \beta_4Income_k +\beta_5Alc_k +\beta_6I*(Drug_k) +  \\ 
     & \quad\: \beta_7I*(Cigar_k) + \beta_cI*(HealthRegion=i) + \epsilon_k
\end{align*}

@@

# Results

>>>>

---
title: "Men and past smokers from ontario are potentially more physically active than women and non-smokers"
subtitle: "Women are engaging in lesser physical activities than men. Furthermore, individuals who smoked before are more likely to engage in physical activity. Consumption of stimulant-related drugs could have also increased as well, resulting in increased overall physical activities in Ontario."
author: 
  - Chris Yong Hong Sen
thanks: "Code and data are available at: [https://github.com/Monoji77/Alcohol_Use_Ontario](https://github.com/Monoji77/Alcohol_Use_Ontario)."
date: today
date-format: long
abstract: "This study explores factors influencing physical activity levels in Ontario, focusing on demographic and substance abuse. Our findings reveal that women engage in less physical activity than men, while individuals who have smoked in the past and those consuming stimulant-related drugs are more likely to be physically active. Additionally, age, highest educational attainment, and health region also play significant roles in determining physical activity levels. These insights highlight the complex interplay of gender, lifestyle behaviors, and sociodemographic factors, providing valuable information for designing targeted public health interventions to promote physical activity."
format: 
  pdf:
    fig-pos: H
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse) # for manipulation and visualisation
library(knitr) # displaying tables
library(arrow) # reading parquet file
library(tmap) # displaying maps
library(lme4) # reporting BIC values
```

@@

Physical activity is a cornerstone of public health, yet participation rates remain suboptimal across various populations in Ontario. This is especially concerning as physical inactivity is a leading contributor to chronic diseases such as heart disease, diabetes, and mental health disorders. Factors such as age, sex, educational attainment, alcohol consumption, smoking habits, and regional disparities in health access play a critical role in shaping physical activity behaviors. Understanding these factors and their complex interactions is crucial for developing targeted interventions that can improve health outcomes and reduce health inequalities in Ontario.

Recent studies suggest that certain demographic groups, such as older adults and individuals with lower educational levels, are more likely to experience barriers to physical activity, while alcohol consumption and smoking are associated with lower physical activity levels [@citationBaumanPhysicalActivity]. However, limited research has explored how these factors specifically interact in Ontario, a province with diverse populations and health regions. As the burden of chronic disease grows, addressing these disparities becomes even more urgent [@citationWarburtonPhysicalActivity] .

This study seeks to address the gaps in understanding by exploring how age, sex, educational attainment, alcohol consumption, smoking, and health regions influence physical activity patterns in Ontario. By examining these relationships, we aim to provide actionable insights that can inform public health policies and interventions tailored to the needs of specific population groups. Our findings will be pivotal in developing strategies to increase physical activity participation and improve health equity in Ontario.

The estimand of this paper would be the actual effect of socio-demographic factors, substance abuse, and health factors on how much time is spent on strenuous activities among individuals in Ontario who engage in physical activities.

The remainder of this paper is structured as follows. @sec-data would provide a sense of the chosen Computing in the Humanities and Social Sciences dataset as well as data exploration of interested variables. @sec-Model goes into the details of the linear regression model fitted on the predictors of interest variables, for which it's results would be showcased in @sec-results. Finally @sec-discussion would interpret the results of the fitted linear regression model, which includes limitations of the paper. At the end of the paper in [Appendix -@sec-diagnostic_checks], this contains diagnostic checks for all models considered.

Overall, this paper was generated using R programming language [@citeR]. Reading of raw data and data manipulation was performed using `Tidyverse` package [@citeTidyverse] and parquet file handling was done using functions in `arrow` package [@citeArrow]. Generation of maps were done using functions in `tmap` package [@citationTmap]. Any other visual plots were using `ggplot`-related functions in Tidyverse [@citeTidyverse] and tables were generated using `kable` function in `knitr` package [@citeKnitR].

@@

Other datasets explored include sources from World Health Organisation and ParticipACTION. World Health Organisation provides aggregated data informing certain indicators of interest such as obesity levels and alcohol consumption. This can be found in the data section for Canada[@OtherSource1]. Also, ParticipACTION provides key statistics regarding proportion of adult population in canada meeting the national guidelines of 150 minutes of moderate to vigorous physical activites, as well as how sedentary canadians are [@OtherSource2]. However, the availability of raw individual level data is absent from these sources. Although the aggregated data gives us a sensing about general Canadian health, we are not able to focus our attention to the Ontario population. The dataset chosen from Canadian Community Health Survey (CCHS) annual component is ideal as it contains individual level data, where we are able to obtain the province the respondents are from.

@@

Our goal is to explore the actual effect of socio-demographic factors, substance abuse, and health factors on how much time is spent on physical activities among active individuals in Ontario. The instrument of measurement would be surveys. Through the cross-sectional survey conducted by CCHS, we are able to get the number of hours spent on physical activities per individual, our response variable.

According to the questionnaire, respondents were asked in a phone survey to indicate whether they had engaged in any strenuous activity that lasted at least 10 minutes and made them sweat and breathe a little harder in the past 7 days. Afterwards, they were asked to indicate the total time spent on those activities. To address the concern of recall bias where an individual may not recall the specific timing in engaging with physical activity, surveyors sectioned physical activity into commuting, sports, work and volunteering. Then within each section, surveyors asked which day (Monday to Sunday) in the past week had they engaged in such activities that made them sweat and breathe a little harder. By doing so, the activities individuals did during the past week would be anchored on the days of the week. Following this would be for the respondents to indicate the total number of hours they engaged in each section of physical activities for the past 7 days.

@@

For socio-demographic factors possibly affecting time spent on physical activites, the surveyors asked the respondent for their age (DHH_AGE) from 0 years old to 121 years old), sex (DHH_SEX) being either male or female, personal income obtained from tax forms submitted to the Canada Revenue Agency upon consent from respondent, geographical health region (ADM_Q037) that was provided by respondents when they agreed to the phone survey, and highest educational attainment (EHG2_04) ranging from less than high school diploma to master's degree and above.

@@

Overall, each entry obtained by respondents assigns quantitative values to each factor and the the response for which we can explore subsequently using linear regression to determine which factors have greater influence in the time spent on physical activities amongst active individuals.

## Response variable: Time spent on physical activity in the past 7 days

From the CCHS dataset, we look at paadvmva variable. This variable is continuous with values ranging from 0 to 9902 representing the number of minutes spent on physical activity in the past 7 days. For easier readability, the variable will be renamed as $time\_spent\_vigorous\_exercise\_7d$.  As explained in section @sec-measurement, this variable measures the time spent on any activities lasting more than 10 minutes that makes an individual sweat more than they normally would. This would include individuals undertaking laborious employment or volunteering efforts, or it could simply include individuals who do cardio and/or strength training.

By using this variable, we are assuming that the routine of an individual from the past 7 days is indicative of their general routine beyond the past 7 days. This variable may not account for individuals who only began a physical routine involving sweat within the past 7 days and were sedentary before, which may not be helpful if included in subsequent model fitting.

Since we are interested in individuals that are already engaging in strenuous activities, we can filter $time\_spent\_vigorous\_exercise\_7d$ to exclude 0. After removing such entries, we obtain the following summary statistics as shown in @tbl-summarised_statistics_time_spent_vigorous_exercise_7d. The median time spent on physical activities the past 7 days would be 300 minutes, which averages out to 43 minutes per day.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summarised_statistics_time_spent_vigorous_exercise_7d
#| tbl-cap: Summarised statistics for time spent on physical activities in the past 7 days

original_data <- read_parquet("../data/02-analysis_data/analysis_data.parquet")

@@

response_var_summarised_data <- data_cleaned_1 |>
  summarise(`Min.` = min(time_spent_vigorous_exercise_7d),
            `1st Qu.` = quantile(time_spent_vigorous_exercise_7d, 0.25),
            Median = median(time_spent_vigorous_exercise_7d),
            `3rd Qu.` = quantile(time_spent_vigorous_exercise_7d, 0.75),
            `Max.` = max(time_spent_vigorous_exercise_7d))

kable(response_var_summarised_data)

@@

Interestingly, the maximum time spent on physical activities the past 7 days was 9902 minutes, and this averages to 1415 minutes per day, which is equivalent to 23.5 hours in a day. This is very unlikely as this suggests that an individual is constantly engaging in physical strenuous activity without sleeping. Logically, this suggests the possibility of outlier values. By observing @fig-boxplot_time_spent_vigorous_exercise_7d, there are many outlier observations for this variable. Roughly 10% of the observed data are considered outliers as they fall outside 1.5 times the interquartile range from the 25% and 75% quartile values. We would be performing log transformation on our response variable. The motivation for this is detailed in [Appendix -@sec-transform-response].

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-boxplot_time_spent_vigorous_exercise_7d
#| fig-cap: Boxplot of time spent on physical activities in the past 7 days

data_cleaned_1 |> 
  ggplot(aes(y=time_spent_vigorous_exercise_7d, x='')) +
  geom_boxplot(outlier.color = 'indianred4', 
               outlier.alpha=0.2,
               width = 0.5) +
  geom_jitter(alpha = 0.01, width=0.15) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  ylab('') +
  xlab('Boxplot of time spent on physical activities in the past 7 days')
```

Subsequently, we would have to remove observations that are very unlikely as they do not represent the general population and could skew the distribution of time spent on physical activities.

@@

```{r response variable exploration}
#| echo: false
#| warning: false
#| message: false
#| label: fig-hist-time_spent_vigorous_exercise_7d
#| fig-cap: Histogram of time spent on physical activities the past 7 days

hist(data_cleaned_1$time_spent_vigorous_exercise_7d,
     main='',
     xlab = '',
     ylab = 'no. of individuals',
     freq=TRUE, cex.main=1, las = 1
    )
```
## Interested explanatory variables {#sec-interested_variables}

There are a total of 13 predictors of interest that is seemingly related to the time spent on physical activities by an individual in Ontario.

@@

From @fig-barplot-age, we can immediately see that most observations fall within age groups 30 to 34 years old (bin 6), and lesser observations fall within age groups 70-74 years old (bin 14).
```{r age predictor barplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot-age
#| fig-cap: Histogram of time spent on physical activites the past 7 days

@@

From @tbl-summarised_statistics_age and @fig-boxplot-age, we are able to see that the the median age bin belongs to age group 40-44 years old. We are able to see that there are no outlier values from @fig-boxplot-age, which means that the survey has obtained an acceptable number of respondents across all age bins. 
```{r age predictor summarised statistics}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summarised_statistics_age
#| tbl-cap: Summarised statistics of respondent ages

age_data_summarised_statistics <- data_cleaned_1 |>
  summarise(`Min.` = min(age),
            `1st Qu.` = quantile(age, 0.25),
            Median = median(age),
            `3rd Qu.` = quantile(age, 0.75),
            `Max.` = max(age))

kable(age_data_summarised_statistics)
```

```{r age predictor boxplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-boxplot-age
#| fig-cap: Boxplot of respondent ages

data_cleaned_1 |>
  ggplot(aes(y=age, x='')) +
  geom_boxplot(outlier.color = 'indianred4', 
               outlier.alpha=0.2) +
  geom_jitter(aes(col = age),
              alpha=0.1, 
              width = 0.15) +
  theme_minimal() +
  scale_y_continuous(breaks = c(0:15)) +
  scale_color_viridis_c(name = 'age bin') +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  ylab('respondent age bin') +
  xlab('')
```

@@

Sex is a possible influential predictor for total time spent on physical activity. Historically, men are associated with exercise and involved in more labour intensive roles such as construction which involves heavy lifting. According to @womenExerciseWHO, women are less active than men by 5% since 2000. However, with the advent of female only establishments such as female gyms, and 32% increase in memberships at fitness and health clubs from 2010 to 2019, it would be interesting to see if the sex of an individual still affects the time spent on physical activites among physically active individuals.

@@

In 2023, @EducationAndActiveLifestyle has done a study to show that highest educational attainment may be a leading factor for physical activity. Therefore, it would be worth exploring the effects of highest levels of education with the time spent on physical activities among the Ontario residents.

@@

```{r proportion of respondents, group(highest_educational_attainment)}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-proportion_highest_educational_attainment
#| tbl-cap: Proportion of respondents grouped by highest educational level

proportion_highest_educ_table <- data_cleaned_1 |>
  count(highest_educational_attainment, name = 'proportion_of_respondents') |>
  mutate(proportion_of_respondents = paste0(round(proportion_of_respondents/nrow(data_cleaned_1)*100,0),'%'),
         highest_educational_attainment = case_when(
           highest_educational_attainment == 1 ~ 'less than sec sch grad',
           highest_educational_attainment == 2 ~ 'sec sch grad',
           highest_educational_attainment == 3 ~ 'post sec certification and above',
           .default = ''))

kable(proportion_highest_educ_table)

@@

Health region is a categorical variable which tells us the geographical location an individual belongs to defined by Provincial Ministries of Health (as detailed by documentation provided by CHASS). It could be possible that certain health regions have more funding that are projected towards increasing awareness towards having an active lifestyle. Therefore, it is worth exploring the relationship between health region and time spent on physical activities.

There are a total of 34 different health regions in the Ontario province. They are encoded using 5 digits. @fig-map_health_region shows us the various health regions in Ontario. The blue regions shows us health regions for which lesser respondents belong to while red regions show us health regions for which more respondents belong to.

@@

By using a GIS focused library in R called tmap [@citationTmap], we are able to generate the map shown in @fig-map_health_region.

```{r health region map}
#| echo: false
#| warning: false
#| message: false
#| label: fig-map_health_region
#| fig-cap: Map of top five and bottom five occuring Ontario health

@@

- 1: No income or income loss
- 2: Less than \$20,000 
- 3: \$20,000 to \$39,999
- 4: \$40,000 to \$59,999
- 5: \$60,000 to \$79,999
- 6: \$80,000 or more

According to @fig-barplot_personal_income, most of our respondents (25% of the sample) in Ontario are earning between \$20,000 to \$39,000 as part of their annual personal income. It is also worth mentioning that there are very few respondents who are not earning any income in that year. They account for less than 0.5% of the sample. In the 2018 annual wages report provided by @PersonalIncomeTaxFiling2018, the annual personal income reported from tax filing in Ontario that year was \$39,510. Therefore, the sampling population's personal income data is representative of Ontario individuals.

@@

### Variable 7: Illicit drug use the past 12 months {#sec-drug_exploration}

This is a binary variable which indicates whether an individual has used drugs in the past 12 months prior to the interview. Drugs include marijuana, hashish, cocaine, amphetamines, speed, methamphetamine, crystal meth, ecstasy, hallucinogens, and sniffing glue, gasoline or other solvents. If they have attempted any of these drugs, this variable would be flagged as '1' and '0' otherwise.

It would be interesting to see if drug consumption has an effect on physical activeness. @PhysicalActivityAndSubstanceAbuse exerts that individuals who are more active tend to have lower substance abuse, suggesting a correlation between these two. Having this variable would be interesting to explore the relationship of an individuals drug usage and their time spent on physical activities.

@@

drug_use_proportion_table <- data_cleaned_1 |>
  count(illicit_drug_use, name = 'num_of_respondents') |>
  mutate(prop_of_population = paste0(round(num_of_respondents/nrow(data_cleaned_1)*100,0),'%'),
         illicit_drug_use = ifelse(illicit_drug_use == 0, 
                                   'never consumed drugs',
                                   'consumed at least once'))

kable(drug_use_proportion_table)
```

@@

It is well known that excessive smoking has negative health impacts on individuals. A cross sectional study conducted by @SmokingAndPhysicalActiveness found that smokers are less likely to be physically active compared to non-smokers. Hence, it would be interesting to discover if there is an effect of smoking more than 100 cigarettes on the time spent on physical activites.

@fig-barplot-cigarettes shows us that within the sample, the visual difference between the number of respondents who smoked more than 100 cigarettes (blue column) and respondents who smoked lesser than 100 cigarettes are not that big. Specifically, 57% of respondents have smoked less than 100 cigarettes while the remaining 43% of respondents smoked more than 100 cigarettes. As the classes in this variable is almost split half,

@@

@fig-pyrmaid-cigarettes shows us cigarette consumption by age bins among out sample population. As respondent age increases from the first bin (18 to 19 year old) to the fourth bin (30 to 34 year old), we can see the number of respondents who have smoked more than 100 cigarettes in their lifetime increased from less than 125 respondents to more than 500 respondents (about 3 times more respondents between the two age groups).
```{r cigarette consumption pyramid figure}
#| echo: false
#| warning: false
#| message: false
#| label: fig-pyrmaid-cigarettes
#| fig-cap: Pyramid of cigarettes consumption in respondent's lifetime by age groups

@@

# Model {#sec-Model}

The goal of our model is to explore which factors affect time spent on physical activities. Therefore a simple exploratory linear regression would be the chosen 
modelling algorithm to determine it. A full model where our response $time\_spent\_vigorous\_exercise\_7d$ would be fitted on all our interested variables outlined in @sec-interested_variables.

@@

\begin{align*} 
\ln(y_k)  &= \beta_0 + \beta_1Age_k + \beta_2Sex_k + \beta_3Educ_k + \beta_4Alc_k +\beta_5I*(Drug_k) + \beta_6I*(Cigar_k) + \\ 
     & \quad\: \beta_7I*(HealthRegion_k=35953) + \beta_8*I(HealthRegion_k=35970) + \epsilon_k
\end{align*}

\begin{align} 
\ln(y_k)|\mu_k, \sigma^2 &\sim \mbox{Normal}(\mu_k, \sigma^2) \tag{1} \\
\mu_k &= \beta_0 + \beta_1Age_k + \beta_2Sex_k + \beta_3Educ_k + \beta_4Alc_k + \tag{2} \\
      & \quad\: \beta_5I*(Drug_k) + \beta_6I*(Cigar_k) + \notag \\ 
      & \quad\: \beta_7I*(HealthRegion_k=35953) + \notag \\
      & \quad\: \beta_8*I(HealthRegion_k=35970) \notag \\
\epsilon_k &\sim \mbox{Normal}(0, \sigma^2) \tag{3}  
\end{align}

@@

### Model justification {#sec-model_justification}

@@

\begin{align*} 
\ln(y_k) &= \beta_0 + \beta_1Age_k + \beta_2Sex_k + \beta_3Educ_k + \beta_4Income_k +\beta_5Alc_k +\beta_6I*(Drug_k) +  \\ 
     & \quad\: \beta_7I*(Cigar_k) + \beta_cI*(HealthRegion=i) + \epsilon_k
\end{align*}

@@

@tbl-BIC_values shows us the BIC values for both the full model and the reduced model. After removing certain explanatory variables, the BIC score has decreased, indicating that the reduced model is has stronger explanatory power while having a less complex model. Therefore, the reduced model is chosen from the original full model.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-BIC_values
#| tbl-cap: Table of BIC values for original and final model

original_model <- 
  readRDS(file = ("../models/original_model.rds"))

# parse in final model
final_model <- readRDS('../models/final_model.rds')
